---
title: "READ.ME"
author: "Max Wegener"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Windows users – installing Rtools (required for compiling R packages)

Some R packages are available only as source code. On Windows you need
Rtools (a collection of compilers and build utilities) so that R can
install those packages. macOS and Linux already ship suitable build
tools; Windows does not.

1)  Do I already have Rtools?

In a fresh R or RStudio session run:

```{r}
Sys.which("make")
```

-   If you get a blank result (""), R cannot find make ⇒ Rtools is
    missing. • If you see a full path containing
    rtools…\usr\bin\make.exe, you’re all set! NO NEED TO GO ANY FURTHER!

2)  Find the Rtools version that matches your R

Check your R version:

```{r}
R.version.string
```

-   Your R version Install this Rtools 4.5.x Rtools45 4.4.x Rtools44
    4.3.x Rtools43 4.2.x Rtools42 (no longer needed if you upgraded R)

3)  Download Rtools

-   Go to <https://cran.r-project.org/bin/windows/Rtools/> Click the
    folder for your version (e.g. rtools43 for 4.3; rtools44 for 4.4;
    rtools45 for 4.5). Download the .exe installer (≈ 200 MB).

4)  Install Rtools

-   Close R / RStudio. Run the downloaded installer. When prompted, tick
    “Add Rtools to system PATH.” Accept the default install path
    (C:\rtools45, C:\rtools44, …). Finish the installation (it may take
    a minute).

5)  Verify the installation

-   After installation open a new R session (or reboot) and run:

```{r}
Sys.which("make")
#>                  make 
#> "C:\\rtools45\\usr\\bin\\make.exe"   # or rtools44 / rtools43
```

#----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

# Mac users – installing gfortran and Xcode Command Line Tools (required for compiling R packages)

-   macOS users – installing gfortran (required for compiling R
    packages)

-   Some R packages require a Fortran compiler to install from source.

-   macOS does not include one by default.

-   You need to install gfortran manually and the version must match
    your R version.

1.  Do I already have gfortran?

In a fresh R or RStudio session run:

```{r}
Sys.which("gfortran") 
```

-   If you get a blank result (""), gfortran is missing ⇒ you need to
    install it.

-   If you see a path like /opt/gfortran/bin/gfortran, you're all set!
    NO NEED TO GO ANY FURTHER!

2.  Find the gfortran version that matches your R Check your R version:

-   Look at the top of the R console window OR

-   copy/paste this into your terminal

```{plaintest}
R.version.string 
```

-   If your R version is 4.5.x Install gfortran

-   If your R version is 4.3 to 4.4.x install gfortran 12.2

-   If you R version is older then no need to install gfortran

3.  Download gfortran

-   <https://mac.r-project.org/tools/gfortran-12.2-universal.pkg>

-   <https://mac.r-project.org/tools/gfortran-14.2-universal.pkg>

4.  Install gfortran

-   Close R / RStudio.

-   Double-click the downloaded .pkg file.

-   Follow the installer prompts and accept the default install path.
    Enter your Mac password when prompted. Finish the installation.

5.  Verify the installation After installation open a new R or RStudio
    session and run:

```{plaintext}
Sys.which("gfortran") 
#\> gfortran 
#\> "/opt/gfortran/bin/gfortran" 
```

-   If you see the path above, gfortran is installed correctly ✅

-   If you still get a blank result, try restarting your Mac and running
    again.

6.  One more check - verify R and gfortran are in sync

-   Run this in R:

```{plaintext}
system("gfortran --version") 
```

You should see output like:

```{plaintext}
GNU Fortran (GCC) 14.2.0 # if you are on R 4.5 
GNU Fortran (GCC) 12.2.0 # if you are on R 4.4 or 4.3 
```

-   If the version number does not match your R version, you have the
    wrong gfortran installed and need to download the correct version
    from <https://mac.r-project.org/tools/>

-   macOS users – fixing missing C++ header files (required for
    compiling R packages)

If you see errors like these when installing packages:

```{plaintext}
fatal error: 'cmath' file not found 
fatal error: 'cstdlib' file not 
found fatal error: 'cstring' file not found 
```

This means your Xcode Command Line Tools are missing or out of date.
This is common after a macOS upgrade.

a.  Do I have this problem?

In R run:

```{plaintext}
system("xcode-select -p")
```

-   If you get an error or a blank result ⇒ Command Line Tools are
    missing. • If you see /Library/Developer/CommandLineTools ⇒ they are
    installed. NO NEED TO GO ANY FURTHER!

b.  Fix the problem Close R and RStudio completely. Open the Mac
    Terminal application and run these commands one at a time:

### Step 1 - Remove the old Command Line Tools:

```{bash}
sudo rm -rf /Library/Developer/CommandLineTools 
```

-   You will be prompted for your Mac password.
-   Type it and press Enter (you will not see the password as you type -
    this is normal).

### Step 2 - Reinstall the Command Line Tools:

```{bash}
xcode-select --install 
```

-   A popup window will appear.
-   Click Install and wait for it to finish.
-   **This may take 5 to 10 minutes depending on your internet
    connection**

### Step 3 - Verify the installation was successful:

```{bash}
xcode-select -p 
```

-   You should see:

```{plaintext}
/Library/Developer/CommandLineTools
```

a.  Try renv again

-   Once the Command Line Tools are installed, reopen RStudio and run:

```{r}
renv::restore() 
```

### ⚠️ Note for macOS beta users:

If you are running a beta version of macOS, package compilation errors
are more likely because R and its tools may not yet be fully compatible
with the newest macOS releases. If you continue to have problems after
reinstalling the Command Line Tools, the most reliable solution is to
use the stable (non-beta) release of macOS.

### 🔑 Quick Reference Card

Your Setup R Version Need

Windows 4.5.x = Rtools45

Windows 4.4.x = Rtools44

Windows 4.3.x = Rtools43

Mac 4.5.x = gfortran 14.2

Mac 4.4.x or 4.3.x = gfortran 12.2

Mac (any version) = any Xcode Command Line Tools
