---
title: "READ.ME"
author: "Max Wegener"
date: "`r Sys.Date()`"
output: html_document
---

Windows users – installing Rtools (required for compiling R packages)

Some R packages are available only as source code.
On Windows you need Rtools (a collection of compilers and build utilities) so that R can install those packages. macOS and Linux already ship suitable build tools; Windows does not.

1)  Do I already have Rtools?

In a fresh R or RStudio session run:

```{r}
Sys.which("make")
```
• If you get a blank result (""), R cannot find make ⇒ Rtools is missing.
• If you see a full path containing rtools…\usr\bin\make.exe, you’re all set! NO NEED TO GO ANY FURTHER!

2)  Find the Rtools version that matches your R

Check your R version:
```{r}
R.version.string
```

Your R version	Install this Rtools
4.5.x	Rtools45
4.4.x	Rtools44
4.3.x	Rtools43
4.2.x	Rtools42 (no longer needed if you upgraded R)

3)  Download Rtools

Go to
https://cran.r-project.org/bin/windows/Rtools/
Click the folder for your version (e.g. rtools43 for 4.3; rtools44 for 4.4; rtools45 for 4.5).
Download the .exe installer (≈ 200 MB).

4)  Install Rtools

Close R / RStudio.
Run the downloaded installer.
When prompted, tick “Add Rtools to system PATH.”
Accept the default install path (C:\rtools45, C:\rtools44, …).
Finish the installation (it may take a minute).

5)  Verify the installation

After installation open a new R session (or reboot) and run:

```{r}
Sys.which("make")
#>                  make 
#> "C:\\rtools45\\usr\\bin\\make.exe"   # or rtools44 / rtools43
```

